### 2017-07-05 paper/introduction.tex
Citations added, small changes

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness plays a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{},
NATURE PAPER, the probabilistic behavior must be considered to faithfully model
the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives rise to probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence, probabilistic programs with nondeterminism have become an active and 
important research focus in program analysis.

@@

\end{compactenum}
While the positive termination implies almost-sure termination, the converse is not true (see~\cite[]{}).

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking functions to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs~\cite{SriramCAV,HolgerPOPL,HONGFEI}, and 
algorithmic 
approaches for special cases such as linear and polynomial RSMs have
also been considered~\cite{POPL,CAV,POPL}.

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{}, for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to this automatically, ranking functions 
of a restricted shape (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{}. 
Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

@@

\item {\em Theoretical foundations.} 
First, we introduce the notion of lexicographic RSMs, and show that
such supermartingales ensure almost-sure termination (\ref{thm:lexrsm-main} in 
Section~\ref{sec:lexicographic}). 
Our first result is a purely mathematical result that introduces a new concept, and 
proves almost-sure termination, that is independent of any probabilistic program.
Based on the mathematical result we show that for probabilistic programs with 
nondeterminism
the existence of a lexicographic RSM with respect to an invariant ensures
almost-sure termination (Theorem~\ref{thm:lexrsm-programs} in 
Section~\ref{sec:lex-programs}).

>>>>

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness plays a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{LearningSurvey,G15},
the probabilistic behavior must be considered to faithfully model the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives rise to probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence, probabilistic programs with nondeterminism have become an active and 
important research focus in program analysis.

@@

\end{compactenum}
While the positive termination implies almost-sure termination, the converse is not true 
(e.g., see Example~\ref{ex:infinite-time}).

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking functions to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{SriramCAV,HolgerPOPL,CF17}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs, and algorithmic approaches for special cases such as 
linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16,CFG16,CNZ17}.

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}, 
for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to this automatically, ranking functions 
of a restricted shape (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

@@

\item {\em Theoretical foundations.} 
First, we introduce the notion of lexicographic RSMs, and show that
such supermartingales ensure almost-sure termination (Theorem~\ref{thm:lexrsm-main} in 
Section~\ref{sec:lexicographic}). 
Our first result is a purely mathematical result that introduces a new concept, and 
proves almost-sure termination, that is independent of any probabilistic program.
Based on the mathematical result we show that for probabilistic programs with 
nondeterminism
the existence of a lexicographic RSM with respect to an invariant ensures
almost-sure termination (Theorem~\ref{thm:lexrsm-programs} in 
Section~\ref{sec:lex-programs}).
